Generalized measurement invariance tests with application to factor analysis

### 2011-12-14 www/paper.Rnw
small touch-ups to R code
(data columns were not correctly formatted, now use StereotypeThreat data
from psychotools-devel)



git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@86 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\documentclass[man]{apa}
\usepackage{graphicx,epsfig,amsmath,alltt,setspace,bm}
%% need no \usepackage{Sweave}
\SweaveOpts{engine = R, eps = TRUE, echo = FALSE, results = hide}

@@

\subsection{Results}
<<mz_mod>>=
load("wicherts05_study1.rda")

d <- subset(d, !is.na(gpa))

## fit model 5b
wdh5b <- lavaan(
   'ability   =~ label(c(rep("load_n", 3), "load_n:min_t")) * numerical + label(rep("load_v", 4)) * verbal + 1 * abstract
    ability   ~  label(c(NA, "lmean:min_c", "lmean:maj_t", "lmean:min_t")) * 1 + c(0, NA, NA, NA) * 1
    ability   ~~ label(c("lvar:maj", "lvar:min", "lvar:maj", "lvar:min")) * ability
    numerical ~  label(c(rep("mean_n", 3), "mean_n:min_t")) * 1
    abstract  ~  label(c("mean_a:maj_c", "mean_a:min_c", rep("mean_a", 2))) * 1
    verbal    ~  label(rep("mean_v", 4)) * 1
    numerical ~~ label(c(rep("var_n", 3), "var_n:min_t")) * numerical
    abstract  ~~ label(rep("var_a", 4)) * abstract
    verbal    ~~ label(rep("var_v", 4)) * verbal',
  data = d, meanstructure = TRUE, group = "group") ##, likelihood = "wishart")

## Refit model with new identification constraints
wdh5b.2 <- lavaan(
   'ability   =~ 1 * verbal + label(c(rep("load_n", 3), "load_n:min_t")) * numerical  + label(rep("load_a", 4)) * abstract
    ability   ~  label(c(NA, "lmean:min_c", "lmean:maj_t", "lmean:min_t")) * 1 + c(0, NA, NA, NA) * 1
    ability   ~~ label(c("lvar:maj", "lvar:min", "lvar:maj", "lvar:min")) * ability
    numerical ~  label(c(rep("mean_n", 3), "mean_n:min_t")) * 1
    abstract  ~  label(c("mean_a:maj_c", "mean_a:min_c", rep("mean_a", 2))) * 1
    verbal    ~  label(rep("mean_v", 4)) * 1
    numerical ~~ label(c(rep("var_n", 3), "var_n:min_t")) * numerical
    abstract  ~~ label(rep("var_a", 4)) * abstract
    verbal    ~~ label(rep("var_v", 4)) * verbal',
  data = d, meanstructure = TRUE, group = "group") #, likelihood = "wishart")
@

@@

\begin{figure}
\caption{Three test statistics of~\eqref{eq:exh0} (set of
  group-specific parameters)
  and~\eqref{eq:exh0.2} (intercept only) across values of student GPA, 
  using data from Study 1 of Wicherts et al.\ (2005).
  Solid, horizontal lines represent critical values
  at $\alpha = 0.05$, and the dotted, horizontal lines (second row)
  represent values of the Cram\'{e}r-von Mises test statistic.}
\label{fig:cusumex}
\setkeys{Gin}{width=\textwidth}
<<gefp-4-1, fig=TRUE, height=8, width=6>>=
## add a small amount of random noise to duplicated GPAs, so that everything
## works with gefp().  (Without this, one can still obtain DM and CvM test 
##                      results but not supLM results.)
set.seed(1090)
ties <- duplicated(d$gpa)
d$gpa[ties] <- runif(sum(ties), min = d$gpa[ties] - .001, max = d$gpa[ties] + .001)

## FIXME It seems like info_full should go in lavaan_S3.R.
info_full <- function(x, ...) solve(vcov(x) * nobs(x))
scores_min_t <- function(x, ...) {
  ef <- estfun(x, ...)
  ef[d$group == "exp:minority", ]
}
gpa_min_t <- subset(d, group == "exp:minority")$gpa

@@

# processes for second identification constraints
gefp_4_info_1 <- gefp(wdh5b.2, fit = NULL, scores = scores_min_t, order.by = gpa_min_t, vcov = info_full, sandwich = FALSE, parm = 15:18)
gefp_1_info_1 <- gefp(wdh5b.2, fit = NULL, scores = scores_min_t, order.by = gpa_min_t, vcov = info_full, sandwich = FALSE, parm = 18)

## visualization
par(mfcol = c(3, 2))
plot(gefp_4_info,  functional = maxBB, main = "DM, k* = 4", xlab = "GPA")
plot(gefp_4_info,  functional = meanL2BB, main = "CvM, k* = 4", xlab = "GPA")
plot(gefp_4_info,  functional = supLM(0.1), main = "max LM, k* = 4", xlab = "GPA")
plot(gefp_1_info, functional = maxBB, main = "DM, k* = 1", xlab = "GPA")
plot(gefp_1_info, functional = meanL2BB, main = "CvM, k* = 1", xlab = "GPA")
plot(gefp_1_info, functional = supLM(0.1), main = "max LM, k* = 1", xlab = "GPA")
@
\end{figure}

>>>>

\documentclass[man]{apa}
\usepackage{graphicx,epsfig,amsmath,alltt,setspace,bm}
%% need no \usepackage{Sweave}
\SweaveOpts{engine = R, eps = FALSE, echo = FALSE, results = hide}

@@

\subsection{Results}
<<wdh_data>>=
## data and transformations
data("StereotypeThreat", package = "psychotools")
## include group variable
StereotypeThreat <- transform(StereotypeThreat, group = interaction(ethnicity, condition))
StereotypeThreat <- StereotypeThreat[order(StereotypeThreat$group),]
## omit NAs in GPA and break ties randomly
StereotypeThreat <- subset(StereotypeThreat, !is.na(gpa))
ties <- duplicated(StereotypeThreat$gpa)
set.seed(1090)
StereotypeThreat$gpa[ties] <- runif(sum(ties),
  min = StereotypeThreat$gpa[ties] - 0.001, max = StereotypeThreat$gpa[ties] + 0.01)
## keep only data with NAs in GPA
wdh <- subset(StereotypeThreat, !is.na(gpa))
@

<<wdh_model>>=
## fit model 5b
wdh5b <- lavaan(
   'ability   =~ label(c(rep("load_n", 3), "load_n:min_t")) * numerical + label(rep("load_v", 4)) * verbal + 1 * abstract
    ability   ~  label(c(NA, "lmean:min_c", "lmean:maj_t", "lmean:min_t")) * 1 + c(0, NA, NA, NA) * 1
    ability   ~~ label(c("lvar:maj", "lvar:min", "lvar:maj", "lvar:min")) * ability
    numerical ~  label(c(rep("mean_n", 3), "mean_n:min_t")) * 1
    abstract  ~  label(c("mean_a:maj_c", "mean_a:min_c", rep("mean_a", 2))) * 1
    verbal    ~  label(rep("mean_v", 4)) * 1
    numerical ~~ label(c(rep("var_n", 3), "var_n:min_t")) * numerical
    abstract  ~~ label(rep("var_a", 4)) * abstract
    verbal    ~~ label(rep("var_v", 4)) * verbal',
  data = wdh, meanstructure = TRUE, group = "group") ##, likelihood = "wishart")

## refit model with new identification constraints
wdh5b.2 <- lavaan(
   'ability   =~ 1 * verbal + label(c(rep("load_n", 3), "load_n:min_t")) * numerical  + label(rep("load_a", 4)) * abstract
    ability   ~  label(c(NA, "lmean:min_c", "lmean:maj_t", "lmean:min_t")) * 1 + c(0, NA, NA, NA) * 1
    ability   ~~ label(c("lvar:maj", "lvar:min", "lvar:maj", "lvar:min")) * ability
    numerical ~  label(c(rep("mean_n", 3), "mean_n:min_t")) * 1
    abstract  ~  label(c("mean_a:maj_c", "mean_a:min_c", rep("mean_a", 2))) * 1
    verbal    ~  label(rep("mean_v", 4)) * 1
    numerical ~~ label(c(rep("var_n", 3), "var_n:min_t")) * numerical
    abstract  ~~ label(rep("var_a", 4)) * abstract
    verbal    ~~ label(rep("var_v", 4)) * verbal',
  data = wdh, meanstructure = TRUE, group = "group") #, likelihood = "wishart")
@

@@

<<gefp-4-1>>=
## FIXME It seems like info_full should go in lavaan_S3.R.
info_full <- function(x, ...) solve(vcov(x) * nobs(x))
scores_min_t <- function(x, ...) {
  ef <- estfun(x, ...)
  ef[wdh$group == "minority.threat", ]
}
gpa_min_t <- subset(wdh, group == "minority.threat")$gpa

@@

# processes for second identification constraints
gefp_4_info_1 <- gefp(wdh5b.2, fit = NULL, scores = scores_min_t, order.by = gpa_min_t, vcov = info_full, sandwich = FALSE, parm = 15:18)
gefp_1_info_1 <- gefp(wdh5b.2, fit = NULL, scores = scores_min_t, order.by = gpa_min_t, vcov = info_full, sandwich = FALSE, parm = 18)
@

\begin{figure}
\caption{Three test statistics of~\eqref{eq:exh0} (set of
  group-specific parameters)
  and~\eqref{eq:exh0.2} (intercept only) across values of student GPA, 
  using data from Study 1 of Wicherts et al.\ (2005).
  Solid, horizontal lines represent critical values
  at $\alpha = 0.05$, and the dotted, horizontal lines (second row)
  represent values of the Cram\'{e}r-von Mises test statistic.}
\label{fig:cusumex}
\setkeys{Gin}{width=\textwidth}
<<gefp-4-1-plot, fig=TRUE, height=8, width=6>>=
par(mfcol = c(3, 2))
plot(gefp_4_info,  functional = maxBB, main = "DM, k* = 4", xlab = "GPA")
plot(gefp_4_info,  functional = meanL2BB, main = "CvM, k* = 4", xlab = "GPA")
plot(gefp_4_info,  functional = supLM(0.1), main = "max LM, k* = 4", xlab = "GPA")
plot(gefp_1_info, functional = maxBB, main = "DM, k* = 1", xlab = "GPA")
plot(gefp_1_info, functional = meanL2BB, main = "CvM, k* = 1", xlab = "GPA")
plot(gefp_1_info, functional = supLM(0.1), main = "max LM, k* = 1", xlab = "GPA")
@
\end{figure}
